% Document Type: LaTeX
% Master File: tutorial.tex
\documentstyle[11pt,relative,alltt,twoside,relative,fancyheadings,boxedminipage,/homes/EHDM/pvs/pvs]{article}
\sloppy

% full list of sections:
%\includeonly{title,intro,informal,pspace,undecide,conclu,ack,rules}

%\pagestyle{myheadings} % page number in upper right corner
%\markboth{Specification and Verification}{}
%\makeindex
\newcommand{\allttinput}[1]{\hozline{\smaller\smaller\smaller\begin{alltt}\input{#1}\end{alltt}}\hozline}
\newenvironment{pvsscript}{\hozline\smaller\smaller\smaller\begin{alltt}}{\end{alltt}\hozline}

\topmargin -10pt
\textheight 8.5in
\textwidth 6.0in
\headheight 15 pt
\columnwidth \textwidth
\oddsidemargin 0.5in
\evensidemargin 0.5in   % fool system for page 0
\setcounter{topnumber}{9}
\renewcommand{\topfraction}{.99}
\setcounter{bottomnumber}{9}
\renewcommand{\bottomfraction}{.99}
\setcounter{totalnumber}{10}
\renewcommand{\textfraction}{.01}
\renewcommand{\floatpagefraction}{.01}
%\newenvironment{smalltt}{\begin{alltt}\small}{\end{alltt}}

\raggedbottom

\font\largett=cmtt10 scaled\magstep2
\font\hugett=cmtt10 scaled\magstep4
\def\opt{{\smaller\sc {\smaller\smaller \&}optional}}
\def\rest{{\smaller\sc {\smaller\smaller \&}rest}}
\def\default#1{[\,{\tt #1}] }
\def\bkt#1{{$\langle$#1$\rangle$}}

\newenvironment{usage}[1]{\item[usage:\hspace*{-0.175in}]#1\begin{description}\setlength{\itemindent}{-0.2in}\setlength{\itemsep}{0.1in}}{\end{description}}

\begin{document}
\pagestyle{empty}
\title{\Large\bf A Tutorial on Specification and Verification Using
PVS\\(Beta Release)}
\author{N.~Shankar and S.~Owre and J.~M.~Rushby\\Computer Science Laboratory\\SRI International\\
Menlo Park CA 94025 USA\\
Phone: (415) 859-5272\\
\{Shankar,Owre,Rushby\}@csl.sri.com}


%\renewcommand{\baselinestretch}{2}
\newenvironment{display}{\begin{alltt}\small\tt\vspace{0.3\baselineskip}}{\vspace{0.3\baselineskip}\end{alltt}}
\newcommand{\choice}{[\!]}
\newcommand{\normtt}[1]{{\obeyspaces {\tt #1 }}}
\newenvironment{pagegroup}{}{}
%\newenvironment{smalltt}{\begin{alltt}\small\tt}{\end{alltt}}
\newenvironment{tdisplay}{\begin{alltt}\footnotesize\tt\vspace{0.3\baselineskip}}{\vspace{0.3\baselineskip}\end{alltt}}
\input{/homes/EHDM/pvs/doc/jmacros}
\input{/homes/EHDM/pvs/doc/lmacros}
\input{/homes/EHDM/pvs/doc/mathprel}
\input{/homes/EHDM/pvs/doc/pvstex}
\def\id#1{\hbox{\tt #1}} %changing ids from roman to tt.
\bibliographystyle{alpha}
%\input{title}
\vspace{4in}
\maketitle
\pagestyle{fancy}
\renewcommand{\sectionmark}[1]{\markboth{{\em #1}}{}\markright{{\em #1}}}
\renewcommand{\subsectionmark}[1]{\markright{\em #1}}
%\lhead[\thepage]{\rightmark}
%\cfoot{\protect\small\bf \fbox{Beta Release}}
%\rhead[\leftmark]{\thepage}
\setcounter{secnumdepth}{2} 
\setcounter{tocdepth}{3}
\begin{abstract}

\newcommand{\PVSlanguage}{S.~Owre, N.~Shankar, and J.~M. Rushby.
{\em The PVS Specification Language (Beta Release)}.
Computer Science Laboratory, SRI International, Menlo Park, CA,
February 1993.}

\newcommand{\PVSuserguide}{S.~Owre, N.~Shankar, and J.~M. Rushby.
{\em User Guide for the PVS Specification and Verification
System (Beta Release)}.   Computer Science Laboratory, SRI
International, Menlo Park, CA, February 1993.}

\newcommand{\PVSprover}{ N.~Shankar, S.~Owre, and J.~M. Rushby.
{\em The PVS Proof Checker: A Reference Manual (Beta Release)}.
Computer Science Laboratory, SRI International, Menlo Park,
CA, February 1993.}

PVS stands for ``Prototype Verification System.'' It consists of a
specification language integrated with support tools and a theorem
prover.  PVS tries to provide the mechanization needed to apply formal
methods both rigorously and productively.

This tutorial serves to introduce PVS.  In the first section, we
briefly sketch the purposes for which PVS is intended and the
rationale behind its design, mention some of the uses that we and
others are making of it, and explain how to get a copy of the system.
In Section 2, we use a simple example to briefly introduce the major
functions of PVS; Sections 3 and 4 then give more detail on the PVS
language and theorem prover, respectively, also using examples.  The
PVS language, system, and theorem prover each have their own reference
manuals,~\footnote{\PVSlanguage}$^{,}$\footnote{\PVSprover}$^{,}$\footnote{\PVSuserguide}
which you will need to study in order to make productive use of the
system.  A pocket reference card, summarizing all the features of the
PVS language, system, and prover is also available.

This tutorial does not introduce the general ideas of formal methods,
nor explain how formal specification and verification can best be
applied to various problem domains; rather, its purpose is to
introduce some of the more unusual and powerful capabilities that are
provided by PVS.  Consequently, this document, and the examples we
use, are somewhat technical and are most suitable for those who
already have some experience with formal methods and wish to
understand how PVS provides mechanized support for some of the more
challenging aspects of formal methods.

\end{abstract}
\pagenumbering{roman}
\newpage
\tableofcontents
\newpage
\pagenumbering{arabic}

\input{overview}
\input{system}
\input{language}
\input{prover}
\input{hardware-eg}

\section{Exercises}

\newtheorem{prob}{Problem}

\begin{prob}
Based on the discussion of the specification of stacks, try to specify a
PVS theory formalizing queues.  Can the PVS abstract datatype facility
be used for specifying queues?
\end{prob}

\begin{prob}
Specify binary trees with value type {\tt T} as a parametric abstract
datatype in PVS.
\end{prob}

\begin{prob}
Specify a PVS theory formalizing {\em ordered\/} binary trees with respect to a
type parameter {\tt T} and a given total-ordering relation, \ie\ define
a predicate {\tt ordered?} that checks if a given binary tree is ordered
with respect to the given total ordering.
\end{prob}

\begin{prob}
  Prove the stack axioms for the definitions stated in {\tt newstacks}.
\end{prob}

\begin{prob}
  Prove the theorems in the theory {\tt half} (Page~\pageref{half}).
\end{prob}

\begin{prob}
  Define the operation for carrying out the ordered insertion of a value
into an ordered binary tree.  Prove that the insertion operator applied
to an ordered binary tree returns an ordered binary tree.
\end{prob}

% \section{References}
{\smaller
%\bibliographystyle{modplain}
\addcontentsline{toc}{section}{References}
\bibliography{mybib,/homes/rushby/jmr,/homes/EHDM/pvs/doc/pvs,/homes/krypton/shankar/tex}
\newpage
\pagestyle{empty}
\mbox{ }
\newpage 
\pagestyle{empty}
\mbox{ }
\newpage 
\end{document}
