% Master File: wift-tutintro.tex
% \documentstyle[11pt,epsf,part,notochead,nobibhead,nfltxsym,relative,cite,alltt,twoside,relative,fancyheadings,boxedminipage,url,/project/pvs/pvs]{article}
\documentclass[11pt,part]{article}
\usepackage{,notochead,nobibhead,relsize,cite,alltt,relsize,fancyheadings,boxedminipage,url}
\sloppy
\newcommand{\allttinput}[1]{\hozline{\smaller\smaller\smaller\begin{alltt}\input{#1}\end{alltt}}\hozline}
\newenvironment{pvsscript}{\hozline\smaller\smaller\smaller\begin{alltt}}{\end{alltt}\hozline}
\topmargin -10pt
\textheight 8.5in
\textwidth 6.0in
\headheight 15 pt
\columnwidth \textwidth
\oddsidemargin 0.5in
\evensidemargin 0.5in   % fool system for page 0
\setcounter{topnumber}{9}
\renewcommand{\topfraction}{.99}
\setcounter{bottomnumber}{9}
\renewcommand{\bottomfraction}{.99}
\setcounter{totalnumber}{10}
\renewcommand{\textfraction}{.01}
\renewcommand{\floatpagefraction}{.01}
\raggedbottom
\font\largett=cmtt10 scaled\magstep2
\font\hugett=cmtt10 scaled\magstep4
\def\opt{{\smaller\sc {\smaller\smaller \&}optional}}
\def\rest{{\smaller\sc {\smaller\smaller \&}rest}}
\def\default#1{[\,{\tt #1}] }
\def\bkt#1{{$\langle$#1$\rangle$}}
\newenvironment{usage}[1]{\item[usage:\hspace*{-0.175in}]#1\begin{description}\setlength{\itemindent}{-0.2in}\setlength{\itemsep}{0.1in}}{\end{description}}
%\renewcommand{\baselinestretch}{2}
\newenvironment{display}{\begin{alltt}\small\tt\vspace{0.3\baselineskip}}{\vspace{0.3\baselineskip}\end{alltt}}
\newcommand{\choice}{[\!]}
\newcommand{\normtt}[1]{{\obeyspaces {\tt #1 }}}
\newenvironment{pagegroup}{}{}
%\newenvironment{smalltt}{\begin{alltt}\small\tt}{\end{alltt}}
\newenvironment{tdisplay}{\begin{alltt}\footnotesize\tt\vspace{0.3\baselineskip}}{\vspace{0.3\baselineskip}\end{alltt}}
%% \input{/project/pvs/doc/jmacros}
%% \input{/project/pvs/doc/lmacros}
%% \input{/project/pvs/doc/mathprel}
%% \input{/project/pvs/doc/pvstex}
\input{jmacros}
\input{lmacros}
\input{mathprel}
\input{pvstex}
%\newlength{\hsbw}
\newenvironment{jmrsession}{\begin{flushleft}
 \setlength{\hsbw}{\linewidth}
 \addtolength{\hsbw}{-\arrayrulewidth}
 \addtolength{\hsbw}{-\tabcolsep}
 \begin{tabular}{@{}|c@{}|@{}}\hline 
 \begin{minipage}[b]{\hsbw}
 \begingroup\small\mbox{ }\\[-1.8\baselineskip]\begin{alltt}}{\end{alltt}\endgroup\end{minipage}\\ \hline 
 \end{tabular}
 \end{flushleft}}
\def\id#1{\hbox{\tt #1}} %changing ids from roman to tt.
\pagestyle{fancy}
\renewcommand{\sectionmark}[1]{\markright{{\em #1}}}
\renewcommand{\subsectionmark}[1]{}
%\lhead[\thepage]{\rightmark}
%\cfoot{\protect\small\bf \fbox{Beta Release}}
%\rhead[\leftmark]{\thepage}
\setcounter{secnumdepth}{2} 
\setcounter{tocdepth}{3}
\setcounter{page}{0} 
\title{{\Huge\bf A Tutorial Introduction to PVS}\\
\smaller  Presented at WIFT '95: Workshop on Industrial-Strength Formal
Specification Techniques, Boca Raton, Florida, April 1995}
\author{Judy Crow, Sam Owre, John Rushby, Natarajan Shankar, Mandayam
Srivas\thanks{Dave Stringer-Calvert provided valuable comments on
earlier versions of this tutorial, and also checked the specifications
and proofs appearing here.   Preparation of this tutorial was
partially funded by NASA Langley Research Center under Contract
NAS1-18969, and by the Advanced Research Projects Agency through NASA
Ames Research Center NASA-NAG-2-891 (Arpa order A721) to
Stanford Unversity.}\\
Computer Science Laboratory\\
SRI International\\
Menlo Park CA 94025 USA
\\ \mbox{ }\\
{\sc www}: {\tt http://www.csl.sri.com/sri-csl-fm.html}
%{\tt Rushby@csl.sri.com}
%\\Phone: +1 (415) 859-5456\ \ Fax: +1 (415) 859-2844
}
\date{Updated June 1995}
\begin{document} 
\maketitle
\begin{abstract}
This document provides an introductory example, a tutorial, and a
compact reference to the PVS verification system.  It is intended to
provide enough information to get you started using PVS, and to help
you appreciate the capabilities of the system and the purposes for
which it is suitable.

% Full documentation for PVS is available in three volumes (language,
% system, and prover), and a number of more advanced tutorial and other
% examples are also available.

\end{abstract}
\thispagestyle{empty}
\newpage
\evensidemargin 0.0in
\mbox{}
\thispagestyle{empty}
\newpage

\pagenumbering{roman}
\setcounter{page}{1} 

%\section*{Table of Contents}
\markboth{}{Contents}

\tableofcontents
\cleardoublepage

\setcounter{page}{0} 
\pagenumbering{arabic}

\section*{Overview}
\addcontentsline{toc}{section}{Overview}
\markboth{}{Overview}

PVS is a {\em verification system\/}:  an interactive environment for
writing formal specifications and checking formal proofs.  It builds on
nearly 20 years experience at SRI in building verification systems, and on
substantial experience with other systems.  The distinguishing feature of
PVS is its synergistic integration of an expressive specification
language and powerful theorem-proving capabilities.  PVS has been
applied successfully to large and difficult applications in both
academic and industrial settings.

PVS provides an expressive specification language that augments
classical higher-order logic with a sophisticated type system
containing predicate subtypes and dependent types, and with
parameterized theories and a mechanism for defining abstract datatypes
such as lists and trees.  The standard PVS types include numbers
(reals, rationals, integers, naturals, and the ordinals to
$\epsilon_{0}$), records, tuples, arrays, functions, sets, sequences,
lists, and trees, etc.  The combination of features in the PVS
type-system is very convenient for specification, but it makes
typechecking undecidable.  The PVS typechecker copes with this
undecidability by generating proof obligations for the PVS theorem
prover.  Most such proof obligations can be discharged automatically.
This liberation from purely algorithmic typechecking allows PVS to
provide relatively simple solutions to issues that are considered
difficult in some other systems (for example, accommodating partial
functions such as division within a logic of total functions), and it
allows PVS to enforce very strong checks on consistency and other
properties (such as preservation of invariants) in an entirely uniform
manner.

PVS has a powerful interactive theorem prover/proof checker.  The
basic deductive steps in PVS are large compared with many other
systems: there are atomic commands for induction, quantifier
reasoning, automatic conditional rewriting, simplification using
arithmetic and equality decision procedures and type information, and
propositional simplification using binary decision diagrams.  The PVS
proof checker manages the proof construction process by prompting the
user for a suitable command for a given subgoal.  The execution of the
given command can either generate further subgoals or complete a
subgoal and move the control over to the next subgoal in a proof.
User-defined proof strategies can be used to enhance the automation in
the proof checker.  Model-checking capabilities used for automatically
verifying temporal properties of finite-state systems have recently
been integrated into PVS\@.  PVS's automation suffices to prove many
straightforward results automatically; for hard proofs, the automation
takes care of the details and frees the user to concentrate on
directing the key steps.

PVS is implemented in Common Lisp---with ancillary functions provided
in C, Tcl/TK, and \LaTeX---and uses GNU Emacs for its interface.
Configured for Sun {\sc Sparc} Workstations running under SunOS
4.1.3,  the system is freely available under license from SRI.

PVS has been used at SRI to undertake proofs of difficult
fault-tolerant
algorithms~\cite{Lincoln&Rushby93:CAV,Lincoln&Rushby93:FTCS,Lincoln&Rushby94:FTP},
to verify the microcode for selected instructions of a complex,
pipelined, commercial microprocessor having 500,000 transistors where
seeded and unseeded errors were found~\cite{Miller&Srivas95}, to
provide an embedding for the Duration Calculus (an interval temporal
logic~\cite{Skakkebaek&Shankar94}), and for several other
applications.  PVS is installed at many sites worldwide, and is in
serious use at about a dozen of them.  There is a growing list of
significant applications undertaken using PVS by people outside SRI.
Many of these can be examined at the WWW site {\tt
http://www.csl.sri.com/sri-csl-fm.html}.

This tutorial is intended to give you an idea of the flavor of PVS, of
the opportunities created by effective mechanization of formal
methods, and an introduction to the use of the system itself.  PVS is
a big and complex system, so we can really only scratch the surface
here.  To make advanced use of the system, you should study the
manuals (there are three volumes: language~\cite{PVS:language},
prover~\cite{PVS:prover}, and system~\cite{PVS:userguide}), and some
of the more substantial applications.

There are three parts to this tutorial.
\begin{itemize}

\item {\em An Introduction to the Mechanized Analysis of Requirements
Specifications Using PVS.} This tutorial introduction shows how PVS
can be used to actively explore and analyze a simple requirements
specification.  It is intended to demonstrate the utility of
mechanized support for formal methods, and the opportunities for
validation and exploration that are created by effective
mechanization.

% \item {\em Formal Verification for Fault-Tolerant Architectures:
% Prolegomena to the Design of PVS}.  This paper from the February 1995
% issue of the IEEE Transactions on Software Engineering describes some
% of the verifications undertaken using PVS and motivates and describes
% some of the design decisions underlying PVS.  It also illustrates
% some of the more advanced language capabilities.

\item {\em Tutorial on Using PVS}.   This introduces many of
the capabilities of PVS by means of simple examples and takes you
through the process of using the system.  While it can be read as an
overview, it is best to have PVS available and to actively follow along.

\item {\em PVS Reference}.    This presents all PVS
system and prover commands, and illustrates the language constructs
in a very compact form.

\end{itemize}

A useful supplement to the material presented here
is~\cite{Owre95:prolegomena}, which describes some of the larger
verifications undertaken using PVS and also motivates and describes
some of the design decisions underlying PVS.

\part[Introduction to Mechanized Analysis of
Specifications Using PVS]{An Introduction to the Mechanized Analysis of
Requirements Specifications Using PVS}
\cleardoublepage
\markboth{Analyzing Specifications Using PVS}{}
\input{phones}
\newpage
\thispagestyle{plain}

%\part{Prolegomena}

\setcounter{section}{0}
\part{Tutorial on Using PVS}
\markboth{Using PVS}{}
\cleardoublepage

\input{overview}
\input{system}
\input{language}
\input{prover}
\input{hardware-eg}

\section{Exercises}

\newtheorem{prob}{Problem}

\begin{prob}
Based on the discussion of the specification of stacks, try to specify a
PVS theory formalizing queues.  Can the PVS abstract datatype facility
be used for specifying queues?
\end{prob}

\begin{prob}
Specify binary trees with value type {\tt T} as a parametric abstract
datatype in PVS.
\end{prob}

\begin{prob}
Specify a PVS theory formalizing {\em ordered\/} binary trees with respect to a
type parameter {\tt T} and a given total-ordering relation, \ie\ define
a predicate {\tt ordered?} that checks if a given binary tree is ordered
with respect to the given total ordering.
\end{prob}

\begin{prob}
  Prove the stack axioms for the definitions stated in {\tt newstacks}.
\end{prob}

\begin{prob}
  Prove the theorems in the theory {\tt half} (Page~\pageref{half}).
\end{prob}

\begin{prob}
  Define the operation for carrying out the ordered insertion of a value
into an ordered binary tree.  Prove that the insertion operator applied
to an ordered binary tree returns an ordered binary tree.
\end{prob}

\setcounter{section}{0}
\part{PVS Reference}
\markboth{PVS Reference}{}
\cleardoublepage
\newcommand{\wiftnewpage}{\newpage}
\newcommand{\refnewpage}{}
\input{refcard}

\newpage
\def\UrlFont{\tt}
\section*{References}
\markboth{}{References}
\bibliographystyle{modalpha}
\addcontentsline{toc}{section}{References}
\bibliography{/homes/rushby/jmr,/project/pvs/doc/pvs,/homes/shankar/tex}


\end{document}


