% Document Type: LaTeX
% Master File: language.tex

\chapter{Declarations}\label{declarations}
\index{declaration|(pidx}

Entities of PVS are introduced by means of \emph{declarations}, which are
the main constituents of PVS specifications.  Declarations are used to
introduce types, variables, constants, formulas, judgements, conversions,
and other entities.  Most declarations have an \emph{identifier} and
belong to a unique theory.  Declarations also have a body which indicates
the \emph{kind} of the declaration and may provide a signature or
definition for the entity.  \emph{Top-level}
declarations\index{declaration!top-level} occur in the formal parameters,
the assertion section and the body of a theory.  \emph{Local}
declarations\index{declaration!local} for variables may be given, in
association with constant and recursive declarations and \emph{binding
expressions} (\eg\ involving \texttt{FORALL} or \texttt{LAMBDA}).
Declarations are ordered within a theory; earlier declarations may not
reference later ones.\footnote{Thus mutual recursion is not directly
supported.  The effect can be achieved with a single recursive function
that has an argument that serves as a switch for selecting between two or
more subexpressions.}

\index{exporting|(}\index{importing|(}
Declarations introduced in one theory may be referenced in another by
means of the \texttt{IMPORTING} and \texttt{EXPORTING} clauses.  The
\texttt{EXPORTING} clause of a theory indicates those entities that may be
referenced from outside the theory.  There is only one such clause for a
given theory.  The \texttt{IMPORTING} clauses provide access to the
entities exported by another theory.  There can be many \texttt{IMPORTING}
clauses in a theory; in general they may appear anywhere a top-level
declaration is allowed.  See Section~\ref{importings} for more details.
\index{importing|)}\index{exporting|)}

PVS allows the overloading\index{overloading} of declaration identifiers.
Thus a theory named \texttt{foo} may declare a constant \texttt{foo} and a
formula \texttt{foo}.  To support this \emph{ad hoc} overloading,
declarations are classified according to kind\index{declaration!kind}; in
PVS the primary kinds are \emph{type}\index{declaration!kind!type},
\emph{prop}\index{declaration!kind!prop},
\emph{expr}\index{declaration!kind!expr}, and
\emph{theory}\index{declaration!kind!theory}.  Type declarations are of
kind \emph{type}, and may be referenced in type declarations, actual
parameters, signatures, and expressions.  Formula declarations are of kind
\emph{prop}, and may be referenced in auto-rewrite declarations
(Section~\ref{auto-rewrite-decls}) or proofs (see the PVS Prover
Guide~\cite{PVS:prover}).  Variable, constant, and recursive definition
declarations are of kind \emph{expr}; these may be referenced in
expressions and actual parameters.  Newly introduced names need only be
unique within a kind, as there is no way, for example, to use an
expression where a type is expected.\footnote{There are a few exceptions,
for example the actual parameters of theories, since theories may be
instantiated with types or expressions.}

\pvsbnf{bnf-decls}{Declarations Syntax}
\index{syntax!declarations}

Declarations generally consist of an
\emph{identifier}\index{declaration!identifier}, an optional list of
\emph{bindings}\index{declaration!binding}, and a
\emph{body}\index{declaration!body}.  The body determines the kind of the
declaration, and the bindings and the body together determine the
signature and definition of the declared entity.  Multiple
declarations\index{declaration!multiple} may be given in compressed form
in which a common body is specified for multiple identifiers; for example
%
\begin{pvsex}
  x, y, z: VAR int
\end{pvsex}
In every case this is treated the same as the expanded form, thus the
above is equivalent to:
\begin{pvsex}
  x: VAR int
  y: VAR int
  z: VAR int
\end{pvsex}

In the rest of this chapter we describe declarations for types, variables,
constants, recursive definitions, macros, inductive and coinductive
definitions, formulas, judgements, conversions, libraries, and
auto-rewrites.  The declarations for theory parameters, importings,
exportings, and theory abbreviations are given in Chapter~\ref{theories}.
Figure~\ref{bnf-decls} gives the syntax for declarations.

\section{Type Declarations}\label{type-declarations}
\index{type declarations|(}

Type declarations are used to introduce new type names to the context.
There are four kinds of type declaration:

\begin{itemize}

\item \emph{uninterpreted type declaration}: \texttt{T:\ TYPE}
\index{uninterpreted type}\index{type!uninterpreted}

\item \emph{uninterpreted subtype declaration}: \texttt{S:\ TYPE FROM T}
\index{uninterpreted subtype}\index{type!uninterpreted subtype}

\item \emph{interpreted type declaration}: \texttt{T:\ TYPE =
int}\index{interpreted type}\index{type!interpreted}

\item \emph{enumeration type declarations}: \texttt{T:\ TYPE = \setb r,
g, b\sete} \index{enumeration types}\index{type!enumeration}

\end{itemize}

These type declarations introduce \emph{type names}\index{type!name}
that may be referenced in type expressions (see Section~\ref{types}).
They are introduced using one of the keywords
\keyword{TYPE}\index{type@\texttt{TYPE}},
\keyword{NONEMPTY\_TYPE}\index{type@\texttt{NONEMPTY\_TYPE}}, or
\keyword{TYPE+}\index{type+@\texttt{TYPE+}}.

\subsection{Uninterpreted Type Declarations}
\index{type!uninterpreted|(}

Uninterpreted types support abstraction by providing a means of
introducing a type with a minimum of assumptions on the type.  An
uninterpreted type imposes almost no constraints on an implementation of
the specification.  The only assumption made on an uninterpreted type
\texttt{T} is that it is disjoint from all other types, except for
subtypes of \texttt{T}.  For example,
\begin{pvsex}
  T1, T2, T3: TYPE
\end{pvsex}
%
introduces three new pairwise disjoint types.  If desired, further
constraints may be put on these types by means of axioms or assumptions
(see Section~\ref{formula-declarations} on
page~\pageref{formula-declarations}).

It should be emphasized that uninterpreted types are important in
providing the right level of abstraction in a specification.  Specifying
the type body may have the undesired effect of restricting the possible
implementations, and cluttering the specification with needless detail.

\index{type!uninterpreted|)}\index{uninterpreted type|)}


\subsection{Uninterpreted Subtype Declarations}
\index{uninterpreted subtype|(}

Uninterpreted subtype declarations are of the form
\begin{pvsex}
  s: TYPE FROM t
\end{pvsex}
\index{FROM@\texttt{FROM}}
This introduces an uninterpreted
\emph{subtype}\index{subtypes}\index{type!subtype} \texttt{s} of
the \emph{supertype}\index{supertype}\index{type!supertype}
\texttt{t}.  This has the same meaning as
\begin{pvsex}
  s_pred: [t -> bool]
  s: TYPE = (s_pred)
\end{pvsex}
%
in which a new predicate is introduced in the first line and the type
\texttt{s} is declared as a \emph{predicate} subtype in the second
line\footnote{This is described in Section~\ref{subtypes}
(page~\pageref{subtypes}).}.  No assumptions are made about uninterpreted
subtypes; in particular, they may or may not be empty, and two different
uninterpreted subtypes of the same supertype may or may not be disjoint.
Of course, if the supertypes themselves are disjoint, then the
uninterpreted subtypes are as well.

\index{uninterpreted subtype|)}

\subsection{Interpreted Type Declarations}
\index{interpreted type declarations|(}\index{type!interpreted|(}

Interpreted type declarations are primarily a means for providing names
for type expressions.  For example,
\begin{pvsex}
  intfun: TYPE = [int -> int]
\end{pvsex}
%
introduces the type name \texttt{intfun} as an abbreviation for the type
of functions with integer domain and range.  Because PVS uses
\emph{structural equivalence}\index{structural equivalence} instead of
\emph{name equivalence}\index{name equivalence}, any type expression
\texttt{T} involving \texttt{intfun} is equivalent to the type expression
obtained by substituting \texttt{[int -> int]} for \texttt{intfun} in
\texttt{T}.  The available type expressions are described in
Chapter~\ref{types} on page~\pageref{types}.

Interpreted type declarations may be given
parameters.\index{parameterized type names} For example, the type of
integer subranges may be given as
\begin{pvsex}
  subrange(m, n: int): TYPE = \setb{}i:int | m <= i AND i <= n\sete
\end{pvsex}
and \texttt{subrange} with two integer parameters may subsequently be used
wherever a type is expected.  Any use of a parameterized type must include
all of the parameters, so currying of the parameters is not allowed.  Note
that \texttt{subrange} may be overloaded to declare a different type in
the same theory without any ambiguity, as long as the number or type of
parameters is different.

\index{type!interpreted|)}\index{interpreted type declarations|)}


\subsection{Enumeration Type Declarations}\label{enum-types}
\index{enumeration types|(}\index{type!enumeration|(}

Enumeration type declarations are of the form
\begin{pvsex}
  enum: TYPE = \setb{}e_1,\ldots, e_n\sete
\end{pvsex}
%
where the \texttt{e\_i} are distinct identifiers which are taken to
completely enumerate the type.  This is actually a shorthand for the
datatype specification
\begin{pvsex}
  enum: DATATYPE
    e_1: e_1?
         \vdots
    e_n: e_n?
  END enum
\end{pvsex}
%
explained in Chapter~\ref{adts}.  Because of this, enumeration types may
only be given as top-level declarations, and are \emph{not} type
expressions.  The advantage of treating them as datatypes is that the
necessary axioms are automatically generated, and the prover has built-in
facilities for handling datatypes.

\index{type!enumeration|)}\index{enumeration types|)}

\index{type declarations|)}


\subsection{Empty versus Nonempty Types}
\label{emptytypes}
\index{nonempty type}
\index{empty type}
\index{type!nonempty|(}\index{type!empty|(}

As noted before, PVS allows empty types, and the term \emph{type} refers
to either empty or nonempty types.  Constants declared to be of a given
type provide elements of the type, so the type must be nonempty or there
is an inconsistency.  Thus whenever a constant is declared, the system
checks whether the type is nonempty, and if it cannot decide that it is
nonempty it generates an \emph{existence TCC}.\index{existence
TCC}\index{TCC!existence} This is the simple explanation, but it is made
somewhat complicated by the considerations of formal parameters,
uninterpreted versus interpreted type declarations, explicit declarations
of nonemptiness, and
\keyword{CONTAINING}\index{CONTAINING@\texttt{CONTAINING}} clauses on type
declarationss, as well as a desire to keep the number of TCCs generated to
a minimum, while guaranteeing soundness.  The details are provided below.

First note that having variables range over an empty type causes no
difficulties,\footnote{If the type \texttt{T} is empty, then the following
two equivalences hold:
\begin{pvsex}
  (FORALL (x: T): p(x)) IFF TRUE \quad \mbox{\textrm{and}} \quad (EXISTS (x: T): p(x)) IFF FALSE
\end{pvsex}}
so variable declarations and variable bindings never trigger the
nonemptiness check.

During typechecking, type declarations may indicate that the type is
nonempty, and constant declarations of a given type require that the type
be nonempty.  When a type is determined to be nonempty, it is marked as
such so that future checks of constants do not trigger more TCCs.  Below
we describe how type declarations are handled first for declarations in the
body of a theory, and then for type declarations that appear in the formal
parameters, as they require special handling.

\paragraph{Theory Body Type Declarations}

\begin{itemize}

\item Uninterpreted type or subtype declarations introduced with the
keyword \keyword{TYPE} may be empty.  Declaring a constant of that type
will lead to a TCC that is unprovable without further axioms.

\item Uninterpreted type declarations introduced with the keyword
\keyword{NONEMPTY\_TYPE}\index{nonempty_type@\keyword{NONEMPTY\_TYPE}}
or \keyword{TYPE+}\index{type+@\texttt{TYPE+}} are assumed to be nonempty.
Thus the type is marked nonempty.

\item Uninterpreted subtype declarations introduced with the keyword
\keyword{NONEMPTY\_TYPE} or \keyword{TYPE+} are assumed to be nonempty, as long as the
supertype is nonempty.  Thus the supertype is checked, and an existence
TCC is generated if the supertype is not known to be nonempty.  Then the
subtype is marked nonempty.

\item The type of an interpreted constant is nonempty, as the definition
provides a witness.

\item Interpreted type declarations introduced with the keyword
\keyword{TYPE} may be nonempty, depending on the type definition.

\item Any interpreted type declaration with a \keyword{CONTAINING} clause
is marked nonempty, and the \keyword{CONTAINING} expression is typechecked
against the specified type.  In this case no existence TCC is generated,
since the \keyword{CONTAINING} expression is a witness to the type.  Of
course, other TCCs may be generated as a result of typechecking the
\keyword{CONTAINING} expression.

\end{itemize}

\paragraph{Formal Type Declarations}

Only uninterpreted (sub)type declarations may appear in the formal
parameters list.

\begin{itemize}

\item Formal type declarations introduced with the \texttt{TYPE} keyword may
be empty.  This is handled according to the occurrences of constant
declarations involving the type.

\item If there is a constant declaration of that type in the formal
parameter list, then no TCCs are generated, since
any instance of the theory will need to provide both the type and a
witness.  The type is marked nonempty in this case.

\item If the type declaration is a formal parameter and a constant is
declared whose type involves the type, but is not the type itself (for
example, if the formal theory parameters are \texttt{[t:\ TYPE, f:\ [t ->
t]]}), then a TCC may be generated, and a comment is added to the TCC
indicating that an assuming clause may be needed in order to discharge the
TCC.  This TCC will be generated only if an earlier constant declaration
hasn't already forced the type to be marked nonempty.  Note that there are
circumstances in which the formal type may be empty but the type
expression involving that type is nonempty.  This is discussed further
below.

\end{itemize}

\subsection{Checking Nonemptiness}\label{nonemptiness-check}
\index{type!nonempty}
The typechecker knows a type to be nonempty under the
following circumstances:
\begin{itemize}

\item The type was declared to be nonempty, using either the
\keyword{NONEMPTY\_TYPE}\index{nonempty_type@\keyword{NONEMPTY\_TYPE}} or
the synonymous \keyword{TYPE+}\index{type+@\texttt{TYPE+}} keyword.  If the
type is uninterpreted, this amounts to an assumption that the type is
nonempty.  If the type has a definition, then an existence TCC is
generated unless the defining type expression is known to be nonempty.

\item The type was declared to have an element using a
\keyword{CONTAINING}\index{CONTAINING@\texttt{CONTAINING}} expression.

\item A constant was declared for the type.  In this case an existence TCC
is generated for the first such constant, after which the type is marked
as nonempty.

\item It was marked as nonempty from an earlier check.

\end{itemize}

Once an unmarked type is determined to be nonempty, it is marked by the
typechecker so that later checks will not generate existence TCCs.  In
addition, the type components are marked as nonempty.  Thus the types that
make up a tuple type, the field types of a record type, and the supertype
of a subtype are all marked.

It is possible for two equivalent types to be marked differently, for
example:
\begin{pvsex}
  t1: TYPE = \setb{}x: int | x > 2\sete
  t2: TYPE = \setb{}x: int | x > 2\sete
  c1: t1
\end{pvsex}
only marks the first type (\texttt{t1}).  Hence, it is best to name your types and
to use those names uniformly.

\index{type!empty|)}
\index{type!nonempty|)}

\section{Variable Declarations}
\index{variables|(}\index{declaration!variables|(}

Variable declarations introduce new variables and associate a type with
them.  These are \emph{logical} variables, not program variables; they
have nothing to do with state---they simply provide a name and associated
type so that binding expressions and formulas can be succinct.
Variables may not be exported.  Variable
declarations also appear in binding expressions such as \texttt{FORALL} and
\texttt{LAMBDA}.  Such local declarations ``shadow'' any earlier
declarations.  For example, in
\begin{pvsex}
  x: VAR bool
  f: FORMULA (FORALL (x: int): (EXISTS (x: nat): p(x)) AND q(x))
\end{pvsex}
%
The occurrence of \texttt{x} as an argument to \texttt{p} is of type
\texttt{nat}, shadowing the one of type \texttt{int}.  Similarly, the
occurrence of \texttt{x} as an argument to \texttt{q} is of type
\texttt{int}, shadowing the one of type \texttt{bool}.

\index{variables|)}\index{declaration!variables|)}

\section{Constant Declarations}\label{constants}
\index{constants|(}\index{declaration!constants|(}

Constant declarations introduce new constants, specifying their type and
optionally providing a value.  Since PVS is a higher order logic, the term
\emph{constant} refers to functions and relations, as well as the usual
(0-ary) constants.  As with types, there are both \emph{uninterpreted} and
\emph{interpreted} \index{constants!interpreted}%
\index{constants!uninterpreted} constants.  Uninterpreted constants make
no assumptions, although they require that the type be nonempty (see
Section~\ref{nonemptiness-check}, page~\pageref{nonemptiness-check}).
Here are some examples of constant declarations:
\begin{pvsex}
  n: int
  c: int = 3
  f: [int -> int] = (lambda (x: int): x + 1)
  g(x: int): int = x + 1
\end{pvsex}
%
The declaration for \texttt{n} simply introduces a new integer constant.
Nothing is known about this constant other than its type, unless further
properties are provided by \texttt{AXIOM}s.  The other three constants are
interpreted.  Each is equivalent to specifying two declarations: \eg\
the third line is equivalent to
\begin{pvsex}
  f: [int -> int]
  f: AXIOM  f = (LAMBDA (x: int): x + 1)
\end{pvsex}
%
except that the definition is guaranteed to form a \emph{conservative
extension}\index{conservative extension} of the theory.  Thus the
theory remains consistent after the declaration is given if it was
consistent before.

The declarations for \texttt{f} and \texttt{g} above are two different ways to
declare the same function.  This extends to more complex arguments, for
example
\begin{pvsex}
  f: [int -> [int, nat -> [int -> int]]] =
     (LAMBDA (x: int): (LAMBDA (y: int), (z: nat): (LAMBDA (w: int):
       x * (y + w) - z)))
\end{pvsex}
%
is equivalent to
\begin{pvsex}
  f(x: int)(y: int, z: nat)(w: int): int = x * (y + w) - z
\end{pvsex}
%
This can be shortened even further if the variables are declared first:
\begin{pvsex}
  x, y, w: VAR int
  z: VAR nat
  f(x)(y,z)(w): int = x * (y + w) - z
\end{pvsex}
%
Finally, a mix of predeclared and locally declared variables is possible:
\begin{pvsex}
  x, y: VAR int
  f(x)(y,(z: nat))(w: int): int = x * (y + w) - z
\end{pvsex}
%
Note the parentheses around \texttt{z:\ nat}; without these, \texttt{y} would
also be treated as if it were declared to be of type \texttt{nat}.

A construct that is frequently encountered when subtypes are involved is
shown by this example
\begin{pvsex}
  f(x: \setb{}x: int | p(x)\sete): int = x + 1
\end{pvsex}
%
There are two useful abbreviations for this expression.  In the first, we
use the fact that the type \texttt{\setb{}x:\ int | p(x)\sete} is equivalent to
the type expression \texttt{(p)} when \texttt{p} has type \texttt{[int ->
bool]}, and we can write
\begin{pvsex}
  f(x: (p)): int = x + 1
\end{pvsex}
%
The second form of abbreviation basically removes the set braces and the
redundant references to the variable, though extra parentheses are
required:
\begin{pvsex}
  f((x: int | p(x))): int = x + 1
\end{pvsex}
%
Which of these forms to use is mostly a matter of taste; in general,
choose the form that is clearest to read for a given declaration.

Note that functions with range type \texttt{bool} are generally referred
to as \emph{predicates}, and can also be regarded as relations or sets.
For example, the set of positive odd numbers can be characterized by a
predicate as follows:
\begin{pvsex}
  odd: [nat -> bool] = (LAMBDA (n: nat): EXISTS (m: nat): n = 2 * m + 1)
\end{pvsex}
%
PVS allows an alternate syntax for predicates that encourages a
set-theoretic interpretation:
\begin{pvsex}
  odd: [nat -> bool] = \setb{}n: nat | EXISTS (m: nat): n = 2 * m + 1\sete
\end{pvsex}

\index{constants|)}

\section{Recursive Definitions}\label{recursive-definitions}
\index{recursive definitions|(}

Recursive definitions are treated as constant declarations, except that
the defining expression is required, and a \emph{measure}\index{measure
function} must be provided, along with an optional well-founded order
relation.\index{well-founded order releation} The same syntax for
arguments is available as for constant declarations; see the preceding
section.

PVS allows a restricted form of recursive definition; mutual
recursion\index{recursion!mutual}\index{mutual recursion} is not allowed,
and the function must be \emph{total},\index{total function} so that the
function is defined for every value of its domain.  In order to ensure
this, recursive functions must be specified with a
\emph{measure}\index{measure}, which is a function whose signature matches
that of the recursive function, but with range type the domain of the
order relation, which defaults to \texttt{<} on \texttt{nat} or
\texttt{ordinal}\index{ordinal}\index{type!ordinal}.  If the order
relation is provided, then it must be a binary relation on the range type
of the measure, and it must be well-founded; a \emph{well-founded} \tcc\
\index{well-founded TCC}\index{TCC!well-founded} is generated if the order
is not declared to be well-founded.

Here is the classic example of the
\texttt{factorial}\index{factorial@\texttt{factorial}} function:
%
\begin{pvsex}
  factorial(x: nat): RECURSIVE nat =
    IF x = 0 THEN 1 ELSE x * factorial(x - 1) ENDIF
    MEASURE (LAMBDA (x: nat): x)
\end{pvsex}
%
The measure is the expression following the \texttt{MEASURE} keyword (the
optional order relation follows a \texttt{BY} keyword after the
measure).  This definition generates a \emph{termination
TCC};\index{TCC!termination}\index{termination TCC} a proof obligation
which must be discharged in order that the function be well-defined.  In
this case the obligation is
%
\begin{pvsex}
  factorial_TCC2: OBLIGATION
    FORALL (x: nat): NOT x = 0 IMPLIES x - 1 < x
\end{pvsex}

It is possible to abbreviate the given \texttt{MEASURE} function by
leaving out the \texttt{LAMBDA} binding.  For example, the measure
function of the factorial definition may be abbreviated to:
\begin{pvsex}
  MEASURE x
\end{pvsex}
The typechecker will automatically insert a lambda binding corresponding
to the arguments to the recursive function if the measure is not already
of the correct type, and will generate a typecheck error if this process
cannot determine an appropriate function from what has been specified.

A termination \tcc\ is generated for each recursive occurrence of the
defined entity within the body of the definition.\footnote{Some of these
may be subsumed by earlier TCCs, and hence will not be displayed with the
\texttt{M-x show-tccs} command.}  It is obtained in one of two ways.  If a
given recursive reference has at least as many arguments provided as
needed by the measure, then the \tcc\ is generated by applying the measure
to the arguments of the recursive call and comparing that to the measure
applied to the original arguments using the order relation.  The
\texttt{factorial} \tcc\ is of this form.  The context of the occurrence
is included in the \tcc; in this case the occurrence is within the
\texttt{ELSE} part of an \texttt{IF-THEN-ELSE} so the negated condition is
an antecedent to the proof obligation.

If the reference does not have enough arguments available, then the
reference is actually given a \emph{recursive signature}\index{recursive
signature} derived from the recursive function as described below.  This
type constrains the domain to satisfy the measure, and the termination
\tcc\ is generated as a \emph{termination-subtype}
\tcc.\index{termination-subtype TCC}\index{TCC!termination-subtype}
Termination-subtype \tccs\ are recognized as such by the occurrence of the
order in the goal of the \tcc.  For example, we could define a
substitution function for terms as follows.
\begin{pvsex}
  term: DATATYPE
  BEGIN
   mk_var(index: nat): var?
   mk_const(index: nat): const?
   mk_apply(fun: term, args: list[term]): apply?
  END term

  subst(x: (var?), y: term)(s: term): RECURSIVE term =
    (CASES s OF
      mk_var(i): (IF index(x) = i THEN y ELSE s ENDIF),
      mk_const(i): s,
      mk_apply(t, ss): mk_apply(subst(x, y)(t), map(subst(x, y))(ss))
     ENDCASES)
  MEASURE s BY <<
\end{pvsex}
Now the first recursive occurrence of \texttt{subst} has all arguments
provided, so the termination TCC is as expected.  The second occurrence
does not have enough arguments.  The recursive signature of that
occurrence is
\begin{pvsex}
  [[(var?), term] -> [\setb{}z1: term | z1 << s\sete -> term]]
\end{pvsex}
Hence the signature of \texttt{subst(x, y)} is \texttt{[\setb{}z1:\ term | z1 <<
s\sete -> term]}, and map is instantiated to \texttt{map[\setb{}z1:\ term | z1 <<
s\sete, term]}, which leads to the TCC
\begin{pvsex}
subst_TCC2: OBLIGATION
  FORALL (ss: list[term], t: term, s: term, x: (var?)):
    s = mk_apply(t, ss) IMPLIES every[term](LAMBDA (z: term): z << s)(ss);
\end{pvsex}
Note that this \texttt{map} instance could be given directly, just don't
make the mistake of providing \texttt{map[term, term]}, as this leads to a
TCC that says every \texttt{term} is \texttt{<<} \texttt{s}.
For the same reason, if the uncurried form of this definition is given,
then a lambda expression will have to be provided and the type will have
to include the measure, for example,
\begin{pvsex}
   subst(x: (var?), y, s: term): RECURSIVE term =
     (CASES s OF
       mk_var(i): (IF index(x) = i THEN y ELSE s ENDIF),
       mk_const(i): s,
       mk_apply(t, ss): mk_apply(subst(x, y, t),
                                 map(LAMBDA (s1: \setb{}z: term|z<<s\sete):
                                       subst(x, y, s1))(ss))
      ENDCASES)
   MEASURE s BY <<
\end{pvsex}

The recursive signature is generated based on the type of the recursive
function and the measure.  For curried functions, it may be that the
measure does not have the entire domain of the recursive function, but
only the first few.  For example, consider the measure for the function
\texttt{f}.
\begin{pvsex}
  f(r: real)(x, y: nat)(b: boolean): RECURSIVE boolean
    = ...
   MEASURE LAMBDA (r: real): LAMBDA (x, y: nat): x
\end{pvsex}
The type of the measure function is \texttt{[real -> [nat, nat -> nat]]},
which is a prefix of the function type.  In deriving the recursive
signature, the last domain type of the measure is constrained (using a
subtype) in the corresponding position of the recursive function type.  In
this case the recursive signature is
\begin{pvsex}
  [real -> [\setb{}z: [nat, nat] | z`1 < x\sete -> [boolean -> boolean]]]
\end{pvsex}
Note that the recursive signature is a dependent type that depends on the
arguments of the recursive function (\texttt{x} in this case), and hence
only applies within the body of the recursive definition.

The formal argument that typechecking the body of a recursive function
using the recursive signature is sound will appear in a future version of
the semantics manual, for now note that simple attempts to subvert this
mechanism do not work, as the following example illustrates.
\begin{pvsex}
  fbad: RECURSIVE [nat -> nat] = fbad
   MEASURE lambda (n: nat): n
\end{pvsex}
This leads an unprovable TCC.
\begin{pvsex}
  fbad_TCC1: OBLIGATION FORALL (x1: nat, x: nat): x < x1;
\end{pvsex}
The TCC results from the comaprison of the expected type \texttt{[nat ->
nat]} to the derived type \texttt{[\setb{}z:\ nat | z < x1\sete -> nat]}.  Remember
that in PVS domains of function types must be equal in order for the
function types to satisfy the subtype relation, and this is exactly what
the TCC states.

\pvstheory{f91-alltt}{Theory \texttt{f91}}{f91-alltt}
\index{f91@{\texttt{f91}}}

When a doubly recursive call is found, the inner recursive calls are
replaced by variables in the termination \tccs\ generated for the outer
calls.  For example, given the theory of Figure~\ref{f91-alltt} the
termination \tcc\ is

\begin{session}
f91_TCC5: OBLIGATION
  FORALL (i: nat,
          v: [i1:
               \setb{}z: nat |
                        (IF z > 101 THEN 0 ELSE 101 - z ENDIF) <
                         (IF i > 101 THEN 0 ELSE 101 - i ENDIF)\sete ->
               \setb{}j: nat | IF i1 > 100 THEN j = i1 - 10 ELSE j = 91 ENDIF\sete]):
    NOT i > 100 IMPLIES
     IF i > 100 THEN v(v(i + 11)) = i - 10 ELSE v(v(i + 11)) = 91 ENDIF;
\end{session}
where the inner calls to \texttt{f91} have been replaced by the
higher-order variable \texttt{v}, with the recursive signature as shown.
Since the obligation forces us to prove the termination condition for all
functions whose type is that of \texttt{f91}, it will also hold for
\texttt{f91}.  This example also illustrates the use of dependent types,
discussed in Section~\ref{dependent-types}.

\pvstheory{ackerman-alltt}{Theory \texttt{ackerman}}{ackerman-alltt}
\index{ackerman@{\texttt{ackerman}}}

In some cases the natural numbers are not a convenient measure; PVS
also provides the \texttt{ordinal}s, which allow recursion with measures up
to $\varepsilon_0$.  This is primarily useful in handling
lexicographical orderings.  For example, in the definition of the
Ackerman function in Figure~\ref{ackerman-alltt},\footnote{There are
ways of specifying \texttt{ackerman} using higher-order functionals, in
which case the measure is again on the natural numbers.} there are two
termination \tccs\ generated (along with a number of subtype \tccs).
The first termination \tcc\ is
\begin{pvsex}
  ack_TCC2:
    OBLIGATION
      (FORALL m, n:
        NOT m = 0 AND n = 0 IMPLIES ackmeas(m - 1, 1) < ackmeas(m, n))
\end{pvsex}
%
and corresponds to the first recursive call of \texttt{ack} in the body of
\texttt{ack}.  In this occurrence, it is known that \texttt{m $\neq$ 0}
and \texttt{n = 0}.  The remaining expression says that the measure
applied to the arguments of the recursive call to \texttt{ack} is less
than the measure applied to the initial arguments of \texttt{ack}.  Note
that the \texttt{<} in this expression is over the \texttt{ordinal}s, not
the \texttt{real}s.

\index{recursive definitions|)}


\section{Macros}\label{macro-declarations}
\index{macros|(}

There are some definitions that are convenient to use, but it's preferable
to have them expanded whenever they are referenced.  To some extent this
can be accomplished using auto-rewrites in the prover, but rewriting is
restricted.  In particular terms in types or actual parameters are not
rewritten; \texttt{typepred} and \texttt{same-name} must be used.  These
both require the terms to be given as arguments, making it difficult to
automate proofs.

The \texttt{MACRO} declaration is used to indicate definitions that are
expanded at typecheck time.  Macro declarations are normal constant
declarations, with the \texttt{MACRO} keyword preceding the
type.\footnote{This is similar to the \texttt{==} form of E\textsc{hdm}.}
For example, after the declaration
\begin{pvsex}
  N: MACRO nat = 100
\end{pvsex}
any reference to \texttt{N} is now automatically replaced by \texttt{100},
including such forms as \texttt{below[N]}.

Macros are not expanded until they have been typechecked.  This is because
the name overloading allowed by PVS precludes expanding during parsing.
TCCs are generated before the definition is expanded.
\index{macros|)}

\input{inductive_defs}

\section{Formula Declarations}\label{formula-declarations}
\index{formula declarations|(}\index{declaration!formulas|(}

Formula declarations introduce \emph{axioms}\index{axioms},
\emph{assumptions}\index{assumptions}, \emph{theorems}\index{theorems},
and \emph{obligations}\index{obligations}.  The identifier associated with
the declaration may be referenced in auto-rewrite declarations (see
Section~\ref{auto-rewrite-decls} and in proofs (see the \texttt{lemma} command
in the PVS Prover Guide~\cite{PVS:prover}).  The expression that makes up
the body of the formula is a boolean expression.  Axioms, assumptions, and
obligations are introduced with the keywords \texttt{AXIOM},
\texttt{ASSUMPTION}, and \texttt{OBLIGATION}, respectively.  Axioms may
also be introduced using the keyword \texttt{POSTULATE}\index{postulate}.
In the prelude postulates are used to indicate axioms that are provable by
the decision procedures, but not from other axioms.  Theorems may be
introduced with any of the keywords
\texttt{CHALLENGE}\index{claim@{\texttt{CHALLENGE}}},
\texttt{CLAIM}\index{claim@{\texttt{CLAIM}}},
\texttt{CONJECTURE}\index{conjecture@{\texttt{CONJECTURE}}},
\texttt{COROLLARY}\index{corollary@{\texttt{COROLLARY}}},
\texttt{FACT}\index{fact@{\texttt{FACT}}},
\texttt{FORMULA}\index{formula@{\texttt{FORMULA}}},
\texttt{LAW}\index{law@{\texttt{LAW}}},
\texttt{LEMMA}\index{lemma@{\texttt{LEMMA}}},
\texttt{PROPOSITION}\index{proposition@{\texttt{PROPOSITION}}},
\texttt{SUBLEMMA}\index{sublemma@{\texttt{SUBLEMMA}}}, or
\texttt{THEOREM}\index{theorem@{\texttt{THEOREM}}}.

Assumptions are only allowed in assuming clauses (see
Section~\ref{assuming}).  Obligations are generated by the system for
\tccs, and cannot be specified by the user.  Axioms are treated
specially when a proof is analyzed, in that they are not expected to
have an associated proof.  Otherwise they are treated exactly like
theorems.  All the keywords associated with theorems have the same
semantics, they are there simply to allow for greater diversity in
classifying formulas.

Formula declarations may contain free variables\index{free variables}, in
which case they are equivalent to the universal closure\index{universal
closure} of the formula.\footnote{The universal closure of a formula is
obtained by surrounding the formula with a \texttt{FORALL} binding
operator whose bindings are the free variables of the formula.  For
example, the universal closure of \texttt{p(x,y) => q(z)} is
\texttt{(FORALL x,y,z:\ p(x,y) => q(z))} (assuming \texttt{x}, \texttt{y}
and \texttt{z} resolve to variables).} In fact, the prover actually uses
the universal closure when it introduces a formula to a proof.  Formula
declarations are the only declarations in which free variables are
allowed.

\index{declaration!formulas|)}\index{formula declarations|)}

\input{judgements}

\input{conversions}

\input{libraries}

\input{auto-rewrite}

\index{declaration|)}
